---
title: "Graphics with ggplot2"
format: html
editor: visual
---

The exercises in this document follow along closely with [Lesson 8](https://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2.html) of [R for Reproducible Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/), provided by the [Software Carpentry](https://software-carpentry.org/) community under the creative commons license, [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/). For more information see [The Carpentries](https://carpentries.org/).

::: callout-tip
## Key Points

-   Use ggplot2 to create plots.
-   Think about graphics in layers: aesthetics, geometry, statistics, scale transformation, and grouping.
:::

```{r}
#| include: false
library(gapminder)
data(gapminder)

library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

# load colorblind friendly color paletts
source('https://tinyurl.com/wjm6ryj')
```

#### Challenge 1

Modify the example so that the figure shows how life expectancy has changed over time:

<!-- Challenge 1 answer begin -->

```{r}
ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp)) + geom_point()
```

<!-- Challenge 1 answer end -->

#### Challenge 2

In the previous examples and challenge we've used the `aes` function to tell the scatterplot **geom** about the **x** and **y** locations of each point. Another *aesthetic* property we can modify is the point *color*. Modify the code from the previous challenge to **color** the points by the "continent" column. What trends do you see in the data? Are they what you expected?

<!-- Challenge 2 answer begin -->

```{r}
# you code goes here
ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp, color = continent)) +
  geom_point()

#this will also work
#ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp)) +
#  geom_point(mapping = aes(color = continent))

#this will color all points the same
#ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp)) + 
#  geom_point(color = "blue")
```

<!-- Challenge 2 answer end -->

#### Challenge 3

Here is a demonstration of drawing points *on top of* lines.

```{r}
ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, by=country, 
                                       color=continent)) +
  geom_line() + 
  geom_point() +
  scale_color_manual(values = cbbPalette)

#setting mapping = aes(color=continent) in geom_line() results in the lines changing colors, but the data points remain unchanged
#ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, by=country)) +
#  geom_line(mapping = aes(color=continent)) + 
#  geom_point() +
#  scale_color_manual(values = cbbPalette)
```

Switch the order of the point and line layers from the previous example. What happened?

<!-- Challenge 3 answer begin -->

```{r}
# Your code goes here
ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, by=country, 
                                       color=continent)) +
  geom_point() + 
  geom_line() +
  scale_color_manual(values = cbbPalette)
```

<!-- Challenge 3 answer end -->

#### Challenge 4a

We can fit a simple relationship to the data by adding another layer, `geom_smooth`:

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
 geom_point(alpha = 0.2) +
 scale_x_log10() +
 geom_smooth(method="lm", size=0.5)

#uses method = "gam" by default due to # of observations in largest group
#ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
#  geom_point(alpha = 0.2) +
#  scale_x_log10() +
#  geom_smooth(formula = y ~ x, size=0.5)

```

Modify the color and size of the points on the point layer in the previous example.

Hint: do not use the `aes` function.

<!-- Challenge 4a answer begin -->

```{r}
# Your code goes here
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  scale_x_log10() +
  geom_point(color = "blue", size = 2) +
  geom_smooth(formula = y ~ x, size = 0.5)
```

<!-- Challenge 4a answer end -->

#### Challenge 4b

Modify your solution above so that the points are now a different shape and are colored by continent with new trendlines. Hint: The **color** argument can be used inside the aesthetic.

<!-- Challenge 4b answer begin -->

```{r}
# Your code goes here
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
  scale_x_log10() +
  geom_point(alpha = 0.5, size = 1, aes(shape = continent)) +
  geom_smooth(formula = y ~ x, linewidth = 0.5) +
  scale_color_manual(values = cbpalette)
  
```

<!-- Challenge 4b answer end -->

#### Challenge 5

1.  Generate boxplots to compare life expectancy between the different continents during the available years.
2.  Rename y axis as Life Expectancy.
3.  Remove x axis labels.

<!-- Challenge 5 answer begin -->

```{r}
 
# ggsave('myfigure1.pdf', fig, width = 7, height = 7, units = 'in', )

fig_1 <- ggplot(data = gapminder, mapping = aes(x = continent, y = lifeExp,
                                       color = continent))+
  geom_boxplot() +
  scale_color_manual(values = cbpalette) +
  scale_y_continuous(limits = c(20,100)) +
  labs(
    x = NULL,
    y = "Life Expectancy",
    title = "Life Expectancy of Various Continents",
    subtitle = "between 1952 and 2007",
    caption = "Gapminder Dataset",
    color = "Continent") +
  facet_wrap(~year) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
fig_1
ggsave('myfigure1.pdf', fig_1, width = 7, height = 7, units = 'in', )
```

<!-- Challenge 5 answer end -->
